INDIGO-DataCloud: a converging evolution of software engineering practices to support European Research e-Infrastructures

### 2019-08-23 paper-jgc.tex
conclusions concluded

%\title{Evolution of Processes in Delivering Reliable Software within the Distributed European Research Infrastructures%\thanks{Grants or other notes
%%about the article that should go on the front page should be
%%placed here. General acknowledgments should be placed at the end of the article.}
%}
\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
%\subtitle{}

@@

The fundamental shift brought by the INDIGO--DC SQA procedures was 
to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
it was packaged within a pre-release stage, the source code itself is verified for each minor change. This strategy 
allows to improve the sustainability of the software, by checking the compliance with code style standards, and its 
reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
significantly lower number of bugs detected by the end users. Consequently, bugs are more effortessly fixed since they 
are early captured and end users interface with more reliable and stable software.

@@

The most noticeable metric that demonstrates a reliability improvement comes from the 
fact that the number of software defects uncovered throughout the pre--release phase 
surpassed by a factor of 30 the ones reported by end users in the EGI e-Infrastructure
during the next two years after the first major release (INDIGO-1). Accordingly, the
sustainability of the software was ...

Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
a progressive quality improvement throughout the lifespan of the project. Indeed,
unit testing coverage dropped for about 18\% of the total products halfway through
the project. This behavior was aligned with the high-demanding periods of software
development, in particular, the previous months before the second major release 
(INDIGO--2). This fact demonstrates that, even considered as best practice, test 
cases are not commonly written right after, or even before, introducing a new unit
or functionality in the code. High coverage values are generally advisable but they
entail time--consuming work that is often neglected by the code owners. The effort 
spent in testing should be oriented to those sections of the code that are more 
likely to cause negative effects to end users. Consequently, a high coverage value 
does not generally provide a linear benefit with respect to the effort spent -- the
SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
it does not indicate whether the key sections of the code are covered. However, unit
testing is a valuable tool to detect defects at coding time, so whenever the test 
cases accompany the changes done in the code, a high unit testing coverage value is 
a good indicator of reliability in the software.
%(related defect increase?, resulting from the INDIGO-2 release?)

% products not developed from scratch challenged the application of SQA policies??

@@

applying the SQA criteria in the development stage. As each change in 
the source code is validated to match the SQA requirements, defects in the software 
are more likely to be detected and fixed at the earliest stage. Consequently, costs
are reduced, since solving bugs at later stages require greater efforts, and the end
user obtains a much more reliable piece of software, as demonstrated by the low rate
of defects found once it is in production. Automation is of paramount importance to 
guarantee the {\sl INDIGO--DataCloud}'s SQA criteria compliance. Thus, through the
application of DevOps practices, a CI/CD scenario was put in place to cover both the
validation and delivery of the {\sl INDIGO--DataCloud} software products. The CI/CD 
pipelines provide the agility to produce releases frequently enough to reduce the 
impact on production systems, since the software gets progressively updated.

In this regard, a pre--production validation,
through the preview testbeds or resource center's early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

The software lifecycle demands as well a continuous measurement and analysis
process, which allows timely reactions to discover and fix defects throughout the
development, release and maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

Lessons learned show that sustainability is a key pursuit. The achievements, in terms of
processes, knowledge, software and infrastructures, shall be profited from former successful
completed projects. In this regard, the more recent project described in this paper, {\sl INDIGO-DataCloud},
may have set the path for future reference in the European software development research ecosystem.
The SQA criteria, describing the quality requirements that might be considered during the entire
software lifecycle, have been published as open access and periodically reviewed to incorporate
new recommendations and best practices. Successful software products created during the lifetime of the
project are maintained leveraging the same source code repositories and CI infrastructure. A seamless
knowledge transfer among subsequent software development initiatives is key for the good progress of
delivering reliable software.

>>>>

%\title{Evolution of Processes in Delivering Reliable Software within the Distributed European Research Infrastructures%\thanks{Grants or other notes
%%about the article that should go on the front page should be
%%placed here. General acknowledgments should be placed at the end of the article.}
%}
%\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
\title{INDIGO-DataCloud: a converging evolution of software engineering practices to support European Research e-Infrastructures}
%\subtitle{}

@@

The fundamental shift brought by the {\sl INDIGO--DataCloud} SQA procedures was 
to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
it was packaged within a pre-release stage, the source code itself is verified for each minor change. This strategy 
allows to improve the sustainability of the software, by checking the compliance with code style standards, and its 
reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
significantly lower number of bugs detected by the end users. Consequently, bugs are more effortessly fixed since they 
are early captured and end users interface with more reliable and stable software.

@@

The most noticeable metric that demonstrates a reliability improvement is represented 
by the ratio between the number of software defects uncovered throughout the pre--release
phase of each of the major releases, {\sl INDIGO--1} and {\sl INDIGO--2}, and the ones 
reported by end users within the EGI e--Infrastructure, where core products were 
deployed. The number of bugs detected in the pre--release phase, filed by the SQA
responsibles of the project, surpassed by a factor of 30 the ones reported by those end 
users, which in absolute terms, this value reached no more than 200 bugs over the
observed timeframe. Considering that these low rate of bugs were uncovered throughout 
the testing phase in the project's preview testbed -- where the new releases were 
deployed in order to validate their integration with other component's production 
versions --, one can infer that the previous phase of code analysis -- running unit and
functional test cases for each minor change -- considerably helped in preventing the 
emergence of a greater quantity of software defects. As it was described, this latter
phase of code analysis included as well the compliance with de--facto style standards, 
improving the readability of the code and, consequently, building on the sustainability
path of the software.

Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
a progressive quality improvement throughout the lifespan of the project. Indeed,
unit testing coverage dropped for about 18\% of the total products halfway through
the project. This behavior was aligned with the high-demanding periods of software
development, in particular, the previous months before the second major release 
(INDIGO--2). This fact demonstrates that, even considered as best practice, test 
cases are not commonly written right after, or even before, introducing a new unit
or functionality in the code. High coverage values are generally advisable but they
entail time--consuming work that is often neglected by the code owners. The effort 
spent in testing should be oriented to those sections of the code that are more 
likely to cause negative effects to end users. Consequently, a high coverage value 
does not generally provide a linear benefit with respect to the effort spent -- the
SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
it does not indicate whether the key sections of the code are covered. However, unit
testing is a valuable tool to detect defects at coding time, so whenever the test 
cases accompany the changes done in the code, a high unit testing coverage value is 
a good indicator of reliability in the software.
% ?? (related defect increase?, resulting from the INDIGO-2 release?)

All along the project we have observed that source code review is the cornerstone to
improve the completeness of the unit test cases, as a commitment to strenghten the 
structural quality of the software. Therefore, we move forward in avoiding the 
aforementioned risks of having high coverage values obtained from meaningless test 
cases, which execute a great amount of statements and lines of the code, but using
incomplete test cases. Code reviewers analyse the content of each change in the 
source code and the suitability -- in the above terms -- of the associated tests, 
being the last checkpoint before the change is added in the production branch. Hence 
the vital importance, as mentioned earlier, of providing the tests alongside the code
for every given change. As we have seen, the production branches of 
{\sl INDIGO--DataCloud} source code repositories were protected against direct pushes,
requiring a manual assessment step before approval through peer code review.
% ?? figure with code reviews or sth about PRs, external participation

Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
software lifecycle through the per--change execution of the functional tests provided 
by the developers and the ultimate stage of integration testing in the aforementioned 
preview testbed. As we have seen, the latter testing strategy was entirely performed 
in a manual way, and appeared as a helpful tool to spot defects after exposing the 
deployed software to interactions with external components. Additionally, the functional 
tests, were primarily useful to check that the software met the design specifications, 
and most importantly, that previously working functionalities are preserved when new 
features are added. As mentioned, these type of testing is meant to be continuously 
executed in an automated manner using the CI infrastructure. However, by the time of
{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked for those 
products that did not provide automated functional tests, tracking the fulfillment of 
their fundamental functional requirements was tougher a change--based approach, 
especially when assessing the regressions. The major issues observed that hindered the
implementation of automated functional tests were releated to the unavailability, at
the time, of libraries and frameworks for performing specific actions -- such as 
testing graphical interfaces -- or due to the code owners' neglect -- insufficient
knowledge of the tools -- or rejection -- time--consuming task --. As a result of this,
those products eventually -- even before {\sl INDIGO--1} -- provided functional test
reports prior to the release, thus being moved away from the CI scenario. We infer
from this experience that automated functional tests should be a goal for modern
software since they complement the structural quality checks by adding the validation
of the software's identified functionalities to improve the effectiveness of the 
DevOps approach.
% figure with % of products with functional tests

As a final remark, we want to highlight the practical experience resulted from the
application of the formerly described software engineering practices to improve the 
validation of a specific application, selected from the existing use cases in the 
{\sl INDIGO--DataCloud} project, thus expanding the scope beyond the core products. 
The outcome was a prototype CI/CD pipeline that analysed the code, packaged and 
tested the application based on a pre--defined set of inputs and reference outputs. 
The pipeline delivered then the new version only if all the stages were successful
completed. Research communities are not usually experts in programming or in related 
software engineering practices, so this fact was a real accomplishment in terms of 
the quality and sustainability of their software.

% ?? products not developed from scratch challenged the application of SQA policies??

@@

%applying the SQA criteria in the development stage. As each change in 
%the source code is validated to match the SQA requirements, defects in the software 
%are more likely to be detected and fixed at the earliest stage. Consequently, costs
%are reduced, since solving bugs at later stages require greater efforts, and the end
%user obtains a much more reliable piece of software, as demonstrated by the low rate
%of defects found once it is in production. Automation is of paramount importance to 
%guarantee the {\sl INDIGO--DataCloud}'s SQA criteria compliance. Thus, through the
%application of DevOps practices, a CI/CD scenario was put in place to cover both the
%validation and delivery of the {\sl INDIGO--DataCloud} software products. The CI/CD 
%pipelines provide the agility to produce releases frequently enough to reduce the 
%impact on production systems, since the software gets progressively updated. 
%
%
%In this regard, a pre--production validation,
%through the preview testbeds or resource center's early adoption -- handled by the
%Staged--rollout process -- alleviates the likelihood of discovering uncovered
%issues at production.

The software lifecycle demands from a continuous measurement and analysis process, 
which allows timely reactions in order to discover and fix defects throughout the
development, release and maintenance stages. Conscientious metric tracking and 
response sits at the core of an appropriate software maturity, as ultimate reliability
can only be achieved by enforcing and empowering the SQA criteria through each stage
of the software lifecycle. Lessons learned show that sustainability is a key pursuit.
The achievements, in terms of processes, knowledge, software and infrastructures, shall
be profited from former successful completed projects. In this regard, 
{\sl INDIGO-DataCloud} may have set a path for future reference in software development
practices for the European research e--Infrastructures. The SQA criteria, describing 
the quality requirements that might be considered during the entire software lifecycle,
have been published as open access and periodically reviewed to incorporate new 
recommendations and best practices. Successful software products created during the
lifetime of the project are maintained leveraging the same source code repositories and
CI infrastructure. A seamless knowledge transfer among subsequent software development
initiatives is key for the good progress of delivering reliable software.
